MLH1 Results Outline

### 2020-02-20 doc/Results.rmd
more slow progress

In order to estimate the genome wide recombination rates across our panel of three subspecies from the house mouse species complex, _M. m. domesticus,_ _M. m. musculus_. and _M. m. molossinus_ using `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` wild derived inbred strains. MLH1 foci were quantified from a total of `r sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "male"))$Ncells)` spermatocytes and `r sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells)` oocytes. The sampling of prophase oocytes is diffecult and was the limiting factor thus there are some strains with only MLH1 counts from spermatocytes, which are included for sex specific analysises.

<!--
transition first aragraph to strain specific
Main points
1. -female strain means are v close  -male strain means are more varied!!
-Comparisons to references / literatures
-house mouse close to minimum of 1  obligate CO per chrm

@@

## female section

@@

#female strain table
#female_mouse.avs_for.analysis
F.strain.table <- ddply(female_mouse.avs_for.analysis, c("strain"), summarise,

@@

@@

lm.female.strain <- glm(mean_co ~ strain, data= female_mouse.avs_for.analysis)
lm.female.strain_sum <- summary(lm.female.strain)

@@

#write.table(lm.female.strain_sum$coefficients, "~./MLH1repo/lm.female.strain.csv", sep=",",row.names = TRUE)

@@

lm.m2.Q12_female <- glm(var ~ strain, data= Q12_female_table)
lm.m2.Q12VAR.female_sum <- summary(lmVAR.female.strain)

@@

### male section

@@

```{r Male.specific.glm, echo=FALSE}
# 1 way anova, is at least 1 strain different from the rest?
# given a lm, how sig is the strain effect
glm.male.strain.subsp <- lm(mean_co ~ subsp * strain, data= just.male_mouse.av)

summry.male.strain.effects <- summary(glm.male.strain.subsp)
round(summry.male.strain.effects$coefficients[,c(1,4)], 5)

@@

#M2
lm.male.strain <- glm(mean_co ~ strain, data= just.male_mouse.av)
lm.male.strain_sum<- summary(lm.male.strain)

@@

the coefficients for glms of the CO means are listed above.

@@

Intial examinations of the sex specific data revealed that.. there is more variety in across the strain means in meals compared to females.

The overall pattern for the female mean MLH1 counts per cell is low variation. 
(the most significantly differnt strains in the female data), For _domesticus^G^_ the G strain is `r G.female.diff` higher the the other means. While _domesticu^LEW^_  is `r round(LEW.female.diff,4)` higher  and _musculus^MSM^_ is `r MSM.female.diff` higher compared to the other strain means in females.

while in the male data the strain average for  _musculus^PWD^_ , _molossinus^MSM^_ and  _musculus^SKIVE^_ are  `r mean(PWD_male$mean_co, na.rm = TRUE) /mean(sans_LewRec_male$mean_co, na.rm = TRUE)`,

`r (mean(MSM_male$mean_co, na.rm = TRUE) / mean(sans_LewRec_male$mean_co, na.rm = TRUE) )`, and `r mean(SKIVE_male$mean_co, na.rm = TRUE) / mean(sans_LewRec_male$mean_co, na.rm = TRUE)` higher than the other strain means respectively.

In the _domesticus_ subspecies the ranges of mouse means are 
`r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "WSB"))$mean_co, na.rm=TRUE  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "WSB"))$mean_co, na.rm=TRUE  )` in _domesticus^WSB^_ , `r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "G"))$mean_co, na.rm=TRUE  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "G"))$mean_co, na.rm=TRUE  )` 
in _domesticus^G^_ `r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "LEW"))$mean_co, na.rm=TRUE  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "LEW"))$mean_co, na.rm=TRUE  )` in _domesticus^LEW^_.

@@

The range is `r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "MSM"))$mean_co, na.rm=TRUE  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "MSM"))$mean_co, na.rm=TRUE  )` in  _molossinus^MSM^_.

-These results are above the minimum of 20 which is the expected minimum following a obligate crossover per chromosomes.

MOVE THESE MODEL RESULTS WITH THE OTHER MODEL RESULTS

Two glm models were run to test the effects of strain and subspecies on mean MLH1 count per cell per mouse. The coefficients for the following model are below.

@@

- There is a low degree of strain varaince in Dom with the range of mouse means `r min( ( male_mouse.av %>% filter(subsp  == "Dom"))$mean_co, na.rm=TRUE  )` to `r max( ( male_mouse.av %>% filter(subsp  == "Dom"))$mean_co, na.rm=TRUE )`

- Musc and Mol have a much larger amount of variance across means with the range in mouse means being `r min( ( male_mouse.av %>% filter(subsp  == "Musc"))$mean_co,na.rm = TRUE  )` to `r max( ( male_mouse.av %>% filter(subsp  == "Musc"))$mean_co,na.rm = TRUE  )` in Musc and `r min( ( male_mouse.av %>% filter(subsp  == "Mol"))$mean_co,na.rm = TRUE  )` to `r max( ( male_mouse.av %>% filter(subsp  == "Mol"))$mean_co, na.rm = TRUE )` and Molossinus.

**general linear models** -- indicate the 3 strains have significant strains effects

Due to this and the significant strain effects they will be designedated into the high rec group for later analyses.

- While there is alot of variance within strains, a general pattern is that PWD, SKIVE and MSM can be classified as 'high rec' strains, there strain averages are

--notes on error / 
-- connections to previous measures in the literature

The same model framework was applied to the male specific data.
The male specific analysis was done to assess the variance across strains. The plots below illustrate mouse level means for MLH1 per cell serperated by subspecies.
<!--
The strain means for females were pretty similar. For female Domesticus strains the means are `r (main.table  %>% filter(sex == "female") %>% filter(subsp == "Dom")  )$mean_co` for _domesticus^WSB^_, _domesticus^G^_, and _domesticus^LEW^_ respectively. In the three Musculus strains the means are `r (main.table  %>% filter(sex == "female") %>% filter(subsp == "Musc")  )$mean_co` for _musculus^PWD^_, _musculus^SKIVE^_, and _musculus^KAZ^_ respectively. In the molossinus strains the means are `r (main.table  %>% filter(sex == "female") %>% filter(subsp == "Mol")  )$mean_co` for _molossinus^MSM^_ and _molossinus^MOLF^_ respectively. 
-->

@@

# Evolutionary framework

In order to test the effects of subspecies, sex and strain on the mean of MLH1 counts per cell, we fit `r length(DF.HetC.MixedModel.HQ$mean_co )` mouse averages of MLH1 foci per cell to a series of mixed and linear models.  (*note on sampling /sample numbers, the numbers of mice across strains, is shown in supplemental table )

Check the DF of the number cited above

<!--

$$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * sex + rand(strain) + \varepsilon $$

This is the break down of mice used within the Mixed model

`r kable(table( DF.HetC.MixedModel.HQ$sex, DF.HetC.MixedModel.HQ$strain))`

@@

```{r MLH1co.lmer.MM, echo=FALSE, warning = FALSE, digits = 3}
#THIS IS THE NEW MIXED MODEL
#should be able to use the loaded 
#DF.HetC.MixedModel.HQ

@@

<!--
 $$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * sex * strain + \varepsilon$$
and 
$$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ sex * strain + \varepsilon$$  -->

```{r MLH1.glmM2, echo=FALSE,  digits = 3}
#strain = fixed
glm.strain.fixed.full <- glm(mean_co ~ subsp * sex * strain, data=DF.HetC.MixedModel.HQ)

summ <- summary(glm.strain.fixed.full)
#table of just coeff anf pvalue
#round(summ$coefficients[,c(1,4)],5 )

@@

Our models were designed to test the effects of the subspecies, sex, their interaction, and strain as a random effect on the mouse means.

All of the effects from the mixed model significant (Table X) 
All of the effects within the mixed model - came out as significant -- (LRT: subsp p= `r MLH1.MMlmer_results$blah`
sex p = `r MLH1.MMlmer_results$blah`, subspecies * sex p = `r MLH1.MMlmer_results$blah`).
(random LTR; strain p= `r MLH1.MMlmer_results$blah`).

@@

strain.fixed <- lm(mean_co ~ sex * strain, data=DF.HetC.MixedModel.HQ)

summ.strain.fixed <- summary(strain.fixed)

@@

Following up with linear models which tested for (subsecies M1), and strain effects alone (M3)
indecated ... identified several signifcant sex by strain interaction effects, (MSM and PWD) (glm; PWD*male p= and MSM-male p = )

In both models strain by sex interaction effects were significant for (p = `r summ.strain.fixed$coefficients[,4][12]` and `r summ.strain.fixed$coefficients[,4][13]` for the second mixed model) for _musculus^PWD^_ and _molossinus^MSM^_  respectively. These results confirm the qualitative observation that within these two strains there has been sex-specific evolution in the average MLH1 count per cell. Additionally, _domesticus^G^_ had significant strain and sex by strain effects (p =`r  summ.strain.fixed$coefficients[,4][3]` and `r summ.strain.fixed$coefficients[,4][10]` for strain and strain by sex effect). These results confirm the qualtative pattern of the _domesticus^G^_ strain having the largest degree of sexual dimorphism (heterochiasmy) in the dataset.

@@

RENAME THE MODELS M2 subsp * strain \* sex

@@

LmerMLH1_M1_Q12_VAR_results <- data.frame(sub.sp = LmerMLH1_M1_Q12_CV_subsp,
sex.results =  LmerMLH1_M1_Q12_CV_sex, interaction.pval = LmerMLH1_M1_Q12_CV_int,  rand.pval = LmerMLH1_M1_Q12_CV_strain)
```
<!--
$$mouse\ variance\ in \ CO \ per cell  ~=~ sex * strain + \varepsilon$$-->

For both the mixed model and linear model, sex was the only significant effect while the subspecies, while interaction and random strain effects were not significant (Table X) (LRT; p= `r round(LmerMLH1_VAR_results$sex.results, 5)`, glm; p= `r round(MLH1var_M3_Q12_sum$coefficients[,4][2], 5)` mixed and glm models respectively).

Since the measures for within mouse variance may be more susceptible to technical error effects from the staining protocol (i.e. background noise). For this reasons, we replicated the model analysis from data sets in the subset of cells with higher quality scores compared to the full data set. (these analysis/ models replicated the results of the full data set (sex being the most significant 
(LTR; p = `r round(LmerMLH1_M1_Q12_VAR_results$sex.results, 5)`, and glm; p = `r  round(MLH1var_M3_Q12_sum$coefficients[2,4], 5)` )

<!--*notes on caveats for calculating variance -- lower power-->
For the estimates of variance, the replicates are at the level of mouse, unlike MLH1 counts, in which the replicates are at the level of the cell. This limits the power for detecting significance differences.

**female strongest effect**, for within mouse variance and coefficient of variance (cv) for both the full and higher quality cell set.

in the sex specific data was analyzed in glm's to test for significance differences in the magnitude of within mouse variance across strains.
The male specific models, showed no significant strain effects

For male specific glms for within mouse variance -- had no significant strain effects
the female data models  --- inconsistant and moderate significance p values

<!--
motivation / background: maybe checkpoint / regulation
limitations: observations at the mouse level, decreased power to estiamate the variance --
setup
-quality / technical error
- variance and cv
The mixed models have most significant variables, 
M3.glm for variance was less significant (0.044181 ), 
the Q12 models in general were more significant
cV's were more significant than var

var, full.mixed, q12,mixed, M3.glm.full, M3.glm.q12
1.618e-10 ***, 2.092e-11, 0.044181 * (male),  0.000228 ***

cv, full.mixed, q12,mixed, M3.glm.full, M3.glm.q12
5.033e-12 ***, 1.315e-12,  3.195e-12 ***, 3.16e-05 ***

@@

For mean CO count, _domesticus^G^_  strain effects in both models (p = )
_musculus^MSM^_ was a significant strain effect on female mean MLH1 counts per cell for only the second model (p= ).

Above are the coefficents for the two glm's of the female specific mouse MLH1 averages, which include subspecies and strain as fixed effects. G has the most consistant significant strain effects in both models. MSM has a pretty low p value in the second model, LEW has a slightly significant pvalue.

MOVE THESE MODEL RESULTS WITH THE OTHER MODEL RESULTS

Two models were fitted to the male specific data to test the effects of strain and subspecies.

# Variation in DSB number per cell

@@

explain why male oer female measures were used, -- mention some measures were made -- but the number of cells was so too small.

In an attempt to locatize the above male specific crossoever number evolution within the meiotic pathway we quantified a marker for DSBs, DMC1, in early prophase spermtaocytes. DMC1 foci were scored from a total of `r sum( (DMC1.mean.table %>%  filter(stage == "L") )$ncells)` leptotene and `r sum( (DMC1.mean.table %>%  filter(stage == "Z") )$ncells)` zygotene staged spermatocytes of (X for each strain) juvenile mice (12 to 18 days) when the proportion of early prophase cells is higher (cite). A subset of strains were choosen to test for differences between the high and low recombination groups; _musculus^PWD^_, _molossinus^MSM^_, _ musculus^KAZ^_ ,  _domesticus^WSB^_ , and _domesticus^G^_. One mouse was sampled for each strain.

(the differences in mean DMC1 foci across the 2 time points confirms expected results. (t.test for pooled DMC1 foci) )

The high recombining groups have significantly more DMC1 for for leptotene (t.test; p=`r ttest.HighLow.L.pval` , one-way-anova; p=`r round(l.anova.pval, 5)`), but not the zygotene cells (t.test; p=`r ttest.HighLow.Z.pval`, one-way-anova; p=`r z.anova.pval`).

The ratios for MLH1:DMC1^Leptotene^ and MLH1:DMC1^zygotene^ are shown in Table X, and can be used as a proxy for the CO:NCO ratio (instead CO:(CO+NCO) ratio).
Notably these ratios do not show a significant difference between the high and the low recombining groups (t.test; p = `r DMC1.Lratio_t.test$p.value` and p=`r DMC1.Zratio_t.test$p.value` for leptotene and zygotene based ratios respectively). These results indicate that the best predictor for CO number may be established before DSB formation (cite Baier et al 2014).

@@

chrm.class.barplot
```

@@

chrm.class.barplot <- ggplot(data = chrm.class.category_MELT, aes(y = mean.percent.type, x = strain, fill=as.factor(variable)) ) + geom_col(position = "dodge")+ggtitle("Chrm Proportions from BivData Data")+ylim(c(0,1)) + facet_wrap(sex~subsp, scales = "free")
#chrm.class.barplot

@@

In order to decompose the cell wide rate, we decided to look at the proportion or chromosomes grouped by the number crossovers on the same bivalent. A total of `r total.biv.all` single bivalents were pooled to calculate the average propotion for each chromosome class (Table X).

**The vast majority of all bivalents in house mouse have one or two crossoveres (1CO an 2CO), even across sexes**

- fits with low rec rates observed in house mouse.

(rates of 0CO, 3CO and 4CO, these are marginal chromsomes classes)

**most of the male proportions are 2/3's 1CO and 1/3 2COs. but The high rec strains are distinguishable by having more 2COs at the cost of 1COs, with closer to 1:1**

**pvales for chi square test, are larger for PSD-MSM compared to low rec strains within subsp, PWD-KAZ and MSM-MOLF**

**these results confirm the intuitive conclusion/result that the rapid male specific evolution in gwRR proceeds though the chromosome level/though increasing the number of 2CO bivalents at the expense of 1CO bivalents**

Six strain have been designated low rec groups, WSB, LEW, G, KAZ, CZECH, and MOLF.
They have significantly different proportions for 1CO and 2CO (chi-square; 1CO p = `r Low_male.1CO.prop.test$p.value` and 2CO p= `r Low_male.2CO.prop.test$p.value`). This significant p value indicates that at least one of the six strains had significant differences).

While the proportions are significantly differennt, PWD male and MSM male proportions are more similar to each other than strains from the same subspcies (PWD and KAZ chi-square test; 1CO p =`r PWD.KAZ.1CO.prop.test$p.value`, 2CO p=`r PWD.KAZ.2CO.prop.test$p.value`)
 and MSM and MOLF chi-square test; 1CO p = `r MSM.MOLF.1CO.prop.test$p.value`, 2CO p=`r MSM.MOLF.2CO.prop.test$p.value`).

@@

- High female strains, G and Lew have significantly more 2CO bivalents -- there are some differences in the female strains.

- The overall male pattern that's most striking is the gradient of 2CO proportions

>>>>

We used counts of MLH1 foci per cell to estimate the genome wide recombination rates across our panel of `r length(unique(  (main.table  %>% filter(species == "M.musculus") )$strain) )` wild derived inbred strains sampled from thress subspecies within the house mouse species complex;  _M. m. domesticus,_ _M. m. musculus_. and _M. m. molossinus_. MLH1 foci were quantified from a total of `r sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "male"))$Ncells)` spermatocytes and `r sum( (main.table %>% filter(species == "M.musculus") %>% filter(sex == "female"))$Ncells)` oocytes. The sampling of prophase oocytes is diffecult and was the limiting factor thus there are some strains with only MLH1 counts from spermatocytes, which are included for sex specific analysises.

<!--
more basic stats, error, comparison to other measures

transition first aragraph to strain specific
Main points
1. -female strain means are v close  -male strain means are more varied!!
-Comparisons to references / literatures
-house mouse close to minimum of 1  obligate CO per chrm

@@

<!-- female section -->

@@

female_mouse.avs_for.analysis <- female_mouse.avs_for.analysis %>% filter(strain != "CAST")
#female strain table
#female_mouse.avs_for.analysis
F.strain.table <- ddply(female_mouse.avs_for.analysis, c("strain"), summarise,

@@

#F.strain.table

@@

MLH1.female.spc.glm.M2 <- glm(mean_co ~ strain, data= female_mouse.avs_for.analysis)
sum_MLH1.female.spc.glm.M2 <- summary(MLH1.female.spc.glm.M2)

@@

#write.table(lm.female.strain_sum$coefficients, "~./MLH1repo/lm.female.strain.csv", sep=",",row.names = TRUE)
```

@@

lm.m2.Q12_female <- glm(var ~ strain, data= Q12_female_table)
lm.m2.Q12VAR.female_sum <- summary(lm.m2.Q12_female)#LEW changes to PWD

@@

<!-- male section -->

@@

```{r Male.specific.glm, echo=FALSE}
# 1 way anova, is at least 1 strain different from the rest?
# given a lm, how sig is the strain effect
glm.male.strain.subsp <- glm(mean_co ~ subsp * strain, data= just.male_mouse.av)

#summry.male.strain.effects <- summary(glm.male.strain.subsp)
#round(summry.male.strain.effects$coefficients[,c(1,4)], 5)

@@

#M2
MLH1.male.specific.glm.M2 <- glm(mean_co ~ strain, data= just.male_mouse.av)
sum_MLH1.male.specific.glm.M2 <- summary(MLH1.male.specific.glm.M2)

@@

@@

Intial examinations of the sex specific data revealed that greater variety in across the strain means in males compared to females. In females, _domesticus^G^_, _musculus^MSM^_, and _domesticu^LEW^_ had the greatest differences in means compared to other strains, `r G.female.diff`, `r round(LEW.female.diff,4)`, and 
`r MSM.female.diff` respectively. While in the male data the strain average for  _musculus^PWD^_ , _molossinus^MSM^_ and  _musculus^SKIVE^_ are  `r mean(PWD_male$mean_co, na.rm = TRUE) /mean(sans_LewRec_male$mean_co, na.rm = TRUE)`, `r (mean(MSM_male$mean_co, na.rm = TRUE) / mean(sans_LewRec_male$mean_co, na.rm = TRUE) )`, and `r mean(SKIVE_male$mean_co, na.rm = TRUE) / mean(sans_LewRec_male$mean_co, na.rm = TRUE)` higher than the other strain means.

<notabley these strains come from Musc and Mol -- Mol natural hybrid of musculus and castaneus. this comments out text?>

The musculus and molossinus subspecies have a much larger amount of variance across means with the range in mouse means being `r min( ( male_mouse.av %>% filter(subsp  == "Musc"))$mean_co,na.rm = TRUE  )` to `r max( ( male_mouse.av %>% filter(subsp  == "Musc"))$mean_co,na.rm = TRUE  )` in Musc and `r min( ( male_mouse.av %>% filter(subsp  == "Mol"))$mean_co,na.rm = TRUE  )` to `r max( ( male_mouse.av %>% filter(subsp  == "Mol"))$mean_co, na.rm = TRUE )` and Molossinus.

<!-- dom, low variation and close to the expected in
within the male means, there is a low degree of strain varaince in _Domesticus_ subspecies with the range of mouse means `r min( ( male_mouse.av %>% filter(subsp  == "Dom"))$mean_co, na.rm=TRUE  )` to `r max( ( male_mouse.av %>% filter(subsp  == "Dom"))$mean_co, na.rm=TRUE )`
-->

<!-- ranges of mouse means
In the _domesticus_ subspecies the ranges of mouse means are 
`r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "WSB"))$mean_co, na.rm=TRUE  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "WSB"))$mean_co, na.rm=TRUE  )` in _domesticus^WSB^_ , `r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "G"))$mean_co, na.rm=TRUE  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "G"))$mean_co, na.rm=TRUE  )` 
in _domesticus^G^_ `r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "LEW"))$mean_co, na.rm=TRUE  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "LEW"))$mean_co, na.rm=TRUE  )` in _domesticus^LEW^_.

@@

The range is `r min( ( female_mouse.avs_for.analysis %>% filter(strain  == "MSM"))$mean_co, na.rm=TRUE  )` to `r max( ( female_mouse.avs_for.analysis %>% filter(strain  == "MSM"))$mean_co, na.rm=TRUE  )` in  _molossinus^MSM^_. 
-->

@@

<!--
The strain means for females were pretty similar. For female Domesticus strains the means are `r (main.table  %>% filter(sex == "female") %>% filter(subsp == "Dom")  )$mean_co` for _domesticus^WSB^_, _domesticus^G^_, and _domesticus^LEW^_ respectively. In the three Musculus strains the means are `r (main.table  %>% filter(sex == "female") %>% filter(subsp == "Musc")  )$mean_co` for _musculus^PWD^_, _musculus^SKIVE^_, and _musculus^KAZ^_ respectively. In the molossinus strains the means are `r (main.table  %>% filter(sex == "female") %>% filter(subsp == "Mol")  )$mean_co` for _molossinus^MSM^_ and _molossinus^MOLF^_ respectively. 
-->

@@

# Evolutionary Framework

<to better understand this varation in RR within an evolutionary framework>

In order to test the effects of subspecies, sex and strain on the mean of MLH1 counts per cell, we fit `r length(DF.HetC.MixedModel.HQ$mean_co )` mouse averages of MLH1 foci per cell to a series of mixed and linear models.

<!--
(*note on sampling /sample numbers, the numbers of mice across strains, is shown in supplemental table )

Check the DF of the number cited above
This is the break down of mice used within the Mixed model
`r kable(table( DF.HetC.MixedModel.HQ$sex, DF.HetC.MixedModel.HQ$strain))`

@@

**M1.Mixed Model** $$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * sex + rand(strain) + \varepsilon $$

Our model, M1., was designed to test the effects of the subspecies, sex, their interaction, and strain as a random effect on the mouse means.

```{r MLH1co.lmer.MM, echo=FALSE, warning = FALSE, digits = 3, message=FALSE}
#THIS IS THE NEW MIXED MODEL
#should be able to use the loaded 
#DF.HetC.MixedModel.HQ

@@

All of the effects within the mixed model were significant (LRT: subsp p= `r MLH1.MMlmer_results$sub.sp`, sex p = `r MLH1.MMlmer_results$sex.results`, subspecies * sex p = `r MLH1.MMlmer_results$interaction.pval`) and random LTR; strain p= `r MLH1.MMlmer_results$rand.pval`).

```{r MLH1.glmM2, echo=FALSE, warning=FALSE,  digits = 3}
#strain = fixed
MLH1.M2.glm.strain.fixed.full <- glm(mean_co ~ subsp * sex * strain, data=DF.HetC.MixedModel.HQ)

summ.MLH1.M2.glm.strain.fixed.full <- summary(MLH1.M2.glm.strain.fixed.full)

#table of just coeff anf pvalue
#round(summ$coefficients[,c(1,4)],5 )

@@

@@

MLH1.M3.strain.fixed <- lm(mean_co ~ sex * strain, data=DF.HetC.MixedModel.HQ)

summ.MLH1.M3.strain.fixed <- summary(MLH1.M3.strain.fixed)

@@

 **M2. Linear Model** $$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subsp * sex * strain + \varepsilon$$

 **M3. LInear Model** $$mouse \ av.\ MLH1\ foci\ per\ cell ~=~ sex * strain + \varepsilon$$

Following up with linear models which tested for (subsecies M2), and strain effects alone (M3)
both showed similar -- results -- **so we just report those from M3. also just report effects which were consisantly significant across both M2 and M3**

<they also have sig strain effects .. PWD strain p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][5]` and MSM strain p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][6]`>

which identified several signifcant sex by strain interaction effects, ( _musculus^PWD^_*male and  _musculus^MSM^_*male ) (glm; p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]`
and p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]`
)

<M2(glm; PWD*male p= `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][14]` and MSM-male p = `r summ.MLH1.M2.glm.strain.fixed.full$coefficients[,4][15]`)>
These results confirm the qualitative observation that within these two strains there has been sex-specific evolution in the average MLH1 count per cell.

<M3 also has a few less significant effects>
<not the reverse coefficient indexing -- this will give the name >

Additionally, _domesticus^G^_ had significant strain and sex by strain effects (p =`r  summ.MLH1.M3.strain.fixed$coefficients[,4][3]` and `r summ.MLH1.M3.strain.fixed$coefficients[,4][10]` for strain and strain by sex effect). These results confirm the qualtative pattern of the _domesticus^G^_ strain having the largest degree of sexual dimorphism (heterochiasmy) in the dataset.

_domesticus^G^_ had significant strain (p= `r summ.MLH1.M3.strain.fixed$coefficients[4,][3]`)
and sex by strain effects (p =`r  summ.MLH1.M3.strain.fixed$coefficients[,4][10]`

<!--
also g strain effect/ dom^G^ strain effect 
`r summ.MLH1.M2.glm.strain.fixed.full$coefficients[12,][4]` 
In both models strain by sex interaction effects were significant for (p = `r summ.MLH1.M3.strain.fixed$coefficients[,4][12]` and `r summ.MLH1.M3.strain.fixed$coefficients[,4][13]` for the second mixed model) for _musculus^PWD^_ and _molossinus^MSM^_  respectively. 
-->
<!-- glms for sex specific data also run -->

A similar glm models (M4) was run on the sex specific data set, (just containing subspecies, strain and the interaction).

 **M4.** $$sex\ specific\ mouse \ av.\ MLH1\ foci\ per\ cell ~=~ subspecies * strain + \varepsilon$$

For the male specific data, three of the strain effects of 13 strains had significant effects on mean MLH1 per cell, PWD (p=`r sum_MLH1.male.specific.glm.M2$coefficients[5,][4]`), SKIVE (p=`r sum_MLH1.male.specific.glm.M2$coefficients[8,][4]`) and MSM (p=`r sum_MLH1.male.specific.glm.M2$coefficients[6,][4]` ). (with effect sizes ranging from 5, 7, and 2 foci respectively). Given these results, theses strains are classified as 'high rec' strains and point to rapid evolution in the recombiantion rate for spermatocytes.

While the female specific linear model has many more had significant strain effects; _domesticus^G^_ (p=`r sum_MLH1.female.spc.glm.M2$coefficients[2,][4]`), _domesticus^LEW^_ (p=`r sum_MLH1.female.spc.glm.M2$coefficients[3,][4]`), _domesticus^PWD^_ (p=`r sum_MLH1.female.spc.glm.M2$coefficients[4,][4]`), _domesticus^MSM^_(p=`r sum_MLH1.female.spc.glm.M2$coefficients[5,][4]`), _domesticus^MOLF^_(p=`r sum_MLH1.female.spc.glm.M2$coefficients[6,][4]`), and _domesticus^KAZ^_(p=`r sum_MLH1.female.spc.glm.M2$coefficients[8,][4]`) the effect sizes have a smaller range of four to one, indicateing specific, but small strain effects for the mean number of MLH1 foci per cell.

@@

Following the same basic framework we used for analyzing mean number of MLH1 foci per cell, we applied a set of mixed and linear models -- to test effects (sex differences for the within mouse variance in MLH1 foci per cell).

 **M5. Mixed Model** $$mouse\ variance\ in \ CO \ per cell ~=~ subsp * sex + rand(strain) + \varepsilon $$

 **M6. Linear Model** $$mouse\ variance\ in \ CO \ per cell ~=~ subsp * sex * strain + \varepsilon$$

 **M7. Linear Model** $$mouse\ variance\ in \ CO \ per cell  ~=~ sex * strain + \varepsilon$$

@@

LmerMLH1_M1_Q12_VAR_results <- data.frame(sub.sp = LmerMLH1_M1_Q12_CV_subsp,
sex.results =  LmerMLH1_M1_Q12_CV_sex, interaction.pval = LmerMLH1_M1_Q12_CV_int,  rand.pval = LmerMLH1_M1_Q12_CV_strain)
```

@@

For both the mixed and linear models, sex was the only significant effect (LRT; p= `r round(LmerMLH1_VAR_results$sex.results, 5)`, glm; p= `r round(MLH1var_M3_Q12_sum$coefficients[,4][2], 5)` mixed and glm models respectively). Since the measures for within mouse variance may be more susceptible to technical error effects from the staining protocol (i.e. background noise). For this reasons, we replicated the model analysis using a subset of cells with higher quality scores. These models replicated the results of the full data set with sex being the most significant effect (LTR; p = `r round(LmerMLH1_M1_Q12_VAR_results$sex.results, 5)`, and glm; p = `r  round(MLH1var_M3_Q12_sum$coefficients[2,4], 5)` ). Sex specific effect linear models indicate that there is no significant difference in the amount of within mouse variance in males. While in females, one strain, _domesticus^LEW^_ has significantly higher within mouse variance (glm; p = `r lmVAR.female.strain_sum$coefficients[3,][4]`) for the model with the full dataset. While in the model using the high quality cells, PWD has the only signicant strain effect (glm; p =`r lm.m2.Q12VAR.female_sum$coefficients[4,][4]`) suggesting a lack of a consistant signal for differences in the amount of within mouse variance in oocytes. For the estimates of variance, the replicates are at the level of mouse, unlike MLH1 counts, in which the replicates are at the level of the cell. This limits the power for detecting significance differences across groups, <but the main/general pattern, is that females have significantly higher within mouse variance compared to males. This general pattern is replicated in higher quality dataset and using the coefficient of variance (cv) as a dependant variable.

<!--
motivation / background: maybe checkpoint / regulation
limitations: observations at the mouse level, decreased power to estiamate the variance --

< SUMMERIZE BOTH THE MLH1 evolution and within mouse variance HERE >

# Variation in DSB Number Per Cell

@@

<!--
explain why male oer female measures were used, -- mention some measures were made -- but the number of cells was so too small.
-->

In an attempt to locatize the above male specific crossover number evolution within the meiotic pathway we quantified a marker for DSBs, DMC1, in early prophase spermtaocytes. DMC1 foci were scored from a total of `r sum( (DMC1.mean.table %>%  filter(stage == "L") )$ncells)` leptotene and `r sum( (DMC1.mean.table %>%  filter(stage == "Z") )$ncells)` zygotene staged spermatocytes form juvenille mice (12 to 18 days). A subset of strains were choosen to test for differences between the high and low recombination groups; _musculus^PWD^_, _molossinus^MSM^_, _musculus^KAZ^_ ,  _domesticus^WSB^_ , and _domesticus^G^_. One mouse was sampled for each strain.

For all strains there were significantly more foci from leptotene cells compared to zygotene (t.test; p = ), which reflects the expected patterns across the timing of meiosis.

Confirming the prediction that (more crossovers are associated with more initial DSBs), the two high recombining strains tested have significantly more DMC1 foci compared to low recombining strains in leptoene cells (t.test; p=`r ttest.HighLow.L.pval` , one-way-anova; p=`r format(l.anova.pval, scientific=F)`).

The differences in DMC1 foci were not significant for zygotene cells between the two groups (t.test; p=`r ttest.HighLow.Z.pval`, one-way-anova; p=`r z.anova.pval`).  <!-- which maybe be attributed to noise effects

The ratios for MLH1:DMC1^Leptotene^ and MLH1:DMC1^zygotene^ are shown in Table X, and can be used as a proxy for the CO:NCO ratio (instead CO:(CO+NCO) ratio). -->

The ratios of average number of MLH1 and DMC1 foci per cell -- have  .... biological significance

The ratios, calculated for DMC1 means from both stages of prophase, are not significantly differet between the strain groups (t.test; p = `r DMC1.Lratio_t.test$p.value` and p=`r DMC1.Zratio_t.test$p.value` for leptotene and zygotene based ratios respectively).

These results indicate that the changes -- which lead to differences in crossover numbers may be established before DSB formation (cite Baier et al 2014).

@@

#chrm.class.barplot
```

@@

chrm.class.barplot <- ggplot(data = chrm.class.category_MELT, aes(y = mean.percent.type, x = strain, fill=as.factor(variable)) ) + geom_col(position = "dodge")+ggtitle("Chrm Proportions from BivData Data")+ylim(c(0,1)) + facet_wrap(sex~subsp, scales = "free")
#chrm.class.barplot
chrm.class.barplot
```

@@

In order to decompose the cell wide rate, we decided to look at the proportion of chromosome classes (ie by the number crossovers on the same bivalent). A total of `r total.biv.all` single bivalents were pooled to calculate the average propotion for each chromosome class (Table X).

The first attern we notice from looking at theis large number of samples is that The vast majority of all bivalents in house mouse have one or two crossoverrs (1CO an 2CO), even across sexes. <these results fit with previous observations on the number of crossovers per chromosome (Stapley 2019)>

< note about the rates of 0CO, 3CO and 4CO? these are marginal chromsomes classes / we'll discuss later?) >

@@

**2. most of the male proportions are 2/3's 1CO and 1/3 2COs. but The high rec strains are distinguishable by having more 2COs at the cost of 1COs, with closer to 1:1**

While the proportions are significantly different, PWD male and MSM male proportions are more similar to each other than strains from the same subspcies (PWD and KAZ chi-square test; 1CO p =`r PWD.KAZ.1CO.prop.test$p.value`, 2CO p=`r PWD.KAZ.2CO.prop.test$p.value`)
 and MSM and MOLF chi-square test; 1CO p = `r MSM.MOLF.1CO.prop.test$p.value`, 2CO p=`r MSM.MOLF.2CO.prop.test$p.value`).
